---
title: "Untitled"
author: "Daniel Schwartz"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(knitr)
library(cowplot)
```

# Goal

To match Refeq phages to hosts. This will enable to assign a phage infecting a sporulator or a non-sporulator host.

```{r load data}
# list of all phages processed by DRAM
d.vir <- read_tsv(here("enrichment","data/Viruses/vMAG_stats.tsv"))

#import virus-host data
vh.db <- read_tsv(here("enrichment","data","virushostdb.tsv") ) 

# #-----------------------------#
# # Downloading virus-host database data
# #-----------------------------#
# # downloaded on 4/May/2021
# library(curl)
# url <- "ftp://ftp.genome.jp/pub/db/virushostdb/virushostdb.tsv"
# vh.db <- read_tsv(curl(url))
# 
# vh.db <- rename_all(vhdb, make.names)
# 
# write_tsv(vh.db, here("data","virushostdb.tsv"))
# #-----------------------------#
```

## match data by taxid

extracting from vh.db a mapping-table between refseq and taxid. There are many viruses that have multiple refseq ids (=multiple sequences), and there are viruses that have multiple hosts (= multiple rows).

```{r}
map_taxid_refseq <- vh.db %>% 
  select(virus.tax.id, refseq.id) %>% 
  # get all sequence IDs
  separate(refseq.id, into = as.character(1:200), sep = ",",fill = "right") %>% 
  pivot_longer(-virus.tax.id, names_to = "num", values_to = "refseq.id") %>% 
  filter(! is.na(refseq.id)) %>% 
  select(-num) %>% 
  # discard replicated ID pairs
  distinct() %>% 
  mutate(refseq.id = trimws(refseq.id))

```


## organize refseq taxID's

```{r}
d.vir <- d.vir %>% 
  # parse viral ID to match vhdb
  mutate(refseq.id = str_remove(refseq.id, "\\..*$")) %>% 
  mutate(refseq.id = trimws(refseq.id))
```

Are we missing any viruses in vh.db?

```{r}
sum(!d.vir$refseq.id %in% map_taxid_refseq$refseq.id)
d.vir$refseq.id[which(!d.vir$refseq.id %in% map_taxid_refseq$refseq.id)]
```

Only 3! We will do without them.

```{r}
#add taxon ID to viral amg data
d.vir <-left_join(d.vir, map_taxid_refseq, by = "refseq.id") 
```

# Dealing with multiple rows(=hosts) per virus in vhdb

* We only need to keep virus-host data that is relavent to refseq viruses
* We will keep only viruses of bacteria (phages)

```{r}
# keep only vh.db data relavent to refseq
vh.db <- semi_join(vh.db, d.vir, by = c("virus.tax.id"))

# keep only phages of bacteria
vh.db <- vh.db %>% 
  filter(str_detect(host.lineage, "^Bacteria"))
```

How much duplication is there?

```{r}
vh.db%>%
  group_by(virus.tax.id, virus.name)%>%
  summarise(N_hosts=n())%>%
  group_by(N_hosts) %>% 
  summarise(n=n()) %>% 
  arrange(desc(N_hosts))
```

Roughly 10% of phages have multiple hosts listed. 

How much variation is there in multiple host taxonomy ?

Typically phages will infect very closely related hosts (sibling strains or species). 

First we will break up the host lineage data so we can check if  hosts are of similar taxonomy. We have made sporulation predictions at the family level so that is the level of interest. However, there is some cleaning to do first, since the the host lineage specification is not uniform in levels specified. 

Some have an intermediate rank between domain and phylum specifically FCB group AND Terrabacteria group. I will remove those for consistency and then separate out the taxonomy.

```{r}
vh.db <- vh.db%>%
  mutate(host.lineage=str_remove(host.lineage," FCB group;"))%>%
  mutate(host.lineage=str_remove(host.lineage," Terrabacteria group;"))

vh.db <- vh.db%>%
  separate(host.lineage, sep="; ",
           into = c("domain", "phylum", "class", "order", "family","genus.etc"), 
           extra = "merge", remove = F)
```

Are there any viruses with hosts of different families?

```{r}
vh.db%>%
  group_by(virus.tax.id, virus.name, family)%>%
  summarise(n=n())%>%
  arrange(desc(n)) %>% 
  group_by(virus.tax.id, virus.name)%>%
  summarise(n=n())%>%
  filter(n>1) %>% 
  arrange(desc(n))

#keep IDs of viruses with multi-family hosts 
multi_fam_id <- vh.db%>%
  group_by(virus.tax.id, virus.name, family)%>%
  summarise(n=n())%>%
  arrange(desc(n)) %>% 
  group_by(virus.tax.id, virus.name)%>%
  summarise(n=n())%>%
  filter(n>1) %>% 
  arrange(desc(n)) %>% 
  pull(virus.tax.id)

```

There are only twelve such viruses. We will get back to those. But first, ee will keep a single host for all the phages with hosts that converge at the family level. Some phages have refseq in evidence for one of their hosts, I think this is the isolation host, so we will prioritize that row.

```{r}
#remove host duplicate
vh.db_clean <- vh.db %>% 
  # remove multi-family, deal with below
  filter(!virus.tax.id %in% multi_fam_id) %>% 
  # mark refseq evidence as frist priority
  mutate(priority1 = grepl("RefSeq", .$evidence)) %>% 
  # count host taxonomy levels, for second priority
  mutate(priority2 = str_count(host.lineage,";")) %>% 
  arrange(desc(priority1), desc(priority2)) %>% 
  # remove duplicated
  mutate(dup = duplicated(virus.tax.id)) %>% 
  filter(!dup) %>% 
  select(-priority1, -priority2, -dup)

# check duplication
duplicated(vh.db_clean$virus.tax.id)%>%sum() 
```

Now we deal with viruses that have more than one family iin host assignment.

```{r}

vh.db%>%
  filter(virus.tax.id %in% multi_fam_id) %>% 
  select(virus.tax.id, virus.name, family)%>%
  distinct() %>%
  arrange(virus.tax.id)
  
```

Most have a family assigned only once and NA in the other. 

For those with a host having NA in family we will keep the other row, with family assigned. 

```{r}
keep_id <- vh.db%>%
  filter(virus.tax.id %in% multi_fam_id) %>% 
  select(virus.tax.id, virus.name, family)%>%
  distinct() %>%
  filter(is.na(family)) %>% 
  pull(virus.tax.id)

# add to clean list
vh.db_clean <- 
vh.db%>%
  filter(virus.tax.id %in% keep_id) %>% 
   distinct() %>% 
  filter(!is.na(family)) %>% 
  # remove duplicated
  mutate(dup = duplicated(virus.tax.id)) %>% 
  filter(!dup) %>% 
  select(-dup) %>% 
  bind_rows(vh.db_clean,.)
 
# check duplication
duplicated(vh.db_clean$virus.tax.id)%>%sum() 
```


What is left?

```{r}
multi_fam_id <- vh.db%>%
  group_by(virus.tax.id, virus.name, family)%>%
  filter(!is.na(family)) %>% 
  summarise(n=n())%>%
  arrange(desc(n)) %>% 
  group_by(virus.tax.id, virus.name)%>%
  summarise(n=n())%>%
  filter(n>1) %>% 
  arrange(desc(n)) %>% 
  pull(virus.tax.id)

vh.db%>%
  filter(virus.tax.id %in% multi_fam_id) %>% 
  select(virus.tax.id, virus.name, phylum,family)%>%
  distinct() %>%
  arrange(virus.tax.id)
```
We are left with four phages, but only one has the potential to infect a sporulator,  *Thermus phage phi OH2*. According to [VHDB](https://www.genome.jp/virushostdb/1352230) the evidence for *Thermaceae* infection is "*Unpublished, from submitter provided text in sequence record AB823818.*", For Geobacillus infection there is a sequenced prophage in the host genome. I will keep the latter. For the other three phages I will discard dupplicates as above.

```{r}
# add to clean list
vh.db_clean <-
vh.db%>%
  filter(virus.tax.id %in% multi_fam_id) %>%   
  # remove Thermus host
  filter(! family =="Thermaceae") %>% 
  # mark refseq evidence as frist priority
  mutate(priority1 = grepl("RefSeq", .$evidence)) %>% 
  # count host taxonomy levels, for second priority
  mutate(priority2 = str_count(host.lineage,";")) %>% 
  arrange(desc(priority1), desc(priority2)) %>% 
  # remove duplicated
  mutate(dup = duplicated(virus.tax.id)) %>% 
  filter(!dup) %>% 
  select(-priority1, -priority2, -dup) %>% 
    bind_rows(vh.db_clean,.)
 
# check duplication
duplicated(vh.db_clean$virus.tax.id)%>%sum() 

```

# add hosts to refseq phages

```{r}

d.vir <- left_join(d.vir, vh.db_clean, by = "virus.tax.id")

write_csv(d.vir, here("enrichment","data/Viruses/refseq_phages_wHost.csv"))
```

