---
title: "Assign host sporulation"
author: "Daniel Schwartz"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(knitr)
library(cowplot)
```

# The Goal

After matching hosts to the viral-Refseq viruses we now assign each viral hosts as a sporulator or non-sporulator.


# import data 
 
* data on hosts of refseq viruses from "A_add-host-DRAM"

* Curated list of sporulation in families of Firmicutes. 



```{r}
#  data on hosts of refseq viruses
d.vir <- read_csv(here("enrichment","data/Viruses/refseq_phages_wHost.csv"))

# Curated list of sporulation in families of Firmicutes
fam_spor <- read_csv(here("gtdb_spor/data/gtdb_families_sporulation.csv")) %>% 
  distinct() %>% 
  mutate(gtdb_dpf = str_c(gtdb_d,gtdb_p, gtdb_f, sep = ";"))
```

This sporulation list was prepared for GTDB taxonomy, but the hosts for resfeq viruses are in NCBI taxonomy.  I will find the GTDB taxonomy for each of the host taxIDs.

# get gtdb data for NCBI taxIDs

Previously I downloaded the metadata of gtdb. I next pull from there all data that has NCBI taxIDs matching assigned hosts of refseq phages

```{r}

# get gtdb by NCBI taxID

f_meta <- list.files(here("gtdb_spor/data/gtdb_downloads"), pattern = "bac120.*tsv")



# filter by ncbi taxa (taxid) present in viral host data
taxIDs <- d.vir$host.tax.id %>%  unique()
taxIDs <- taxIDs[!is.na(taxIDs)]
# remove a general bacterial term (used for unclassified bacteria)
taxIDs <- taxIDs[taxIDs!=1869227]
# filter function
# function to filter firmicutes by chunks
f <- function(x, pos) {
  x %>% filter(ncbi_taxid %in% taxIDs)
  }

d_meta <-
  read_tsv_chunked(here("gtdb_spor/data/gtdb_downloads", f_meta),
                   DataFrameCallback$new(f))

d_meta <-
  d_meta %>% 
  select(gtdb_taxonomy, ncbi_taxid) %>% 
  separate(gtdb_taxonomy, sep = ";", into = paste0("gtdb_",c("d","p","c","o","f","g","s"))) %>% 
  select(ncbi_taxid, paste0("gtdb_",c("d","p","c","o","f")))

d_meta %>% 
  select(ncbi_taxid, gtdb_f) %>% 
  distinct() %>% 
  group_by(ncbi_taxid) %>% 
  summarise(n_fam=n()) %>% 
  arrange(desc(n_fam))

```

some NCBI taxiIDs map on to multiple GTDB families. I will inspect.

```{r}
# Get tax IDs by number of assigned GTDB families
taxid_fam_count <- d_meta %>% 
  select(ncbi_taxid, gtdb_f) %>% 
  distinct() %>% 
  right_join(., tibble(ncbi_taxid = taxIDs))

taxid_absent <- 
  taxid_fam_count %>% 
  filter(is.na(gtdb_f)) %>% 
  pull(ncbi_taxid)

taxid_unique_fam <- 
  taxid_fam_count %>% 
  filter(!is.na(gtdb_f)) %>% 
  group_by(ncbi_taxid) %>% 
  summarise(n=n()) %>% 
  filter(n==1) %>% 
  pull(ncbi_taxid)
  
taxid_multi_fam <- 
  taxid_fam_count %>% 
  filter(!is.na(gtdb_f)) %>% 
  group_by(ncbi_taxid) %>% 
  summarise(n=n()) %>% 
  filter(n > 1) %>% 
  pull(ncbi_taxid)

```

## TaxIDs mapping to unique familiy

```{r}
d_meta_clean <- 
  d_meta %>% 
  filter(ncbi_taxid %in% taxid_unique_fam) %>% 
  distinct()

# check for duplication
anyDuplicated(d_meta_clean$ncbi_taxid)
```


## TaxIDs mapping to multiple families

From insection it seems that in many cases the majority are consistent in family asignment. See plot:

```{r}

d_meta %>% 
  filter(ncbi_taxid %in% taxid_multi_fam) %>% 
  group_by_all() %>% 
  summarise(n=n()) %>% 
  arrange(n) %>% 
  group_by(ncbi_taxid) %>% 
  mutate(fam_x = LETTERS[row_number()]) %>% 
  mutate(ncbi_taxid = ncbi_taxid %>% as_factor()) %>% 
  ggplot(aes(ncbi_taxid, n)) +
  geom_col(aes(fill = fam_x), show.legend = T)+
  facet_wrap(~ncbi_taxid, scales = "free")+
  theme_classic()+
  scale_fill_viridis_d(direction = -1)+
  theme(  strip.background = element_blank(),
          strip.text.x = element_blank())
```

The only real exxeption is taxid=1409, which is *unclassified Bacillus*. In the next plot I test if the different family assigments differ in sporulation assignment.

```{r}
d_meta %>% 
  filter(ncbi_taxid %in% taxid_multi_fam) %>% 
  group_by_all() %>% 
  summarise(n=n()) %>% 
  arrange(n) %>% 
  left_join(., select(fam_spor, gtdb_f, f_spor)) %>% 
  mutate(ncbi_taxid = ncbi_taxid %>% as_factor()) %>% 
  ggplot(aes(ncbi_taxid, n)) +
  geom_col(aes(fill = f_spor, color = gtdb_f), show.legend = T, color = "red")+
  facet_wrap(~ncbi_taxid, scales = "free")+
  theme_classic()+
  scale_fill_viridis_d(direction = )+
  theme(  strip.background = element_blank(),
          strip.text.x = element_blank())
```

The majority is consistent in asigning sporulation status. In taxIDs 1502 and 1509 each have over 100 rows (=genomes) assigned as sporulators, and a single row assigned as a non-sporulator. I will follow the majority.

```{r}
d_meta_clean <- 
  d_meta %>% 
  filter(ncbi_taxid %in% taxid_multi_fam) %>% 
  group_by_all() %>% 
  summarise(n=n()) %>% 
  group_by(ncbi_taxid) %>% 
  slice_max(n) %>% 
  select (-n) %>% 
  bind_rows(d_meta_clean,.)

# check for duplication
anyDuplicated(d_meta_clean$ncbi_taxid)
```
At this point we can assign sporulation status using the gtdb family.

```{r}
d_meta_clean <-
  fam_spor %>% 
  select(gtdb_p,gtdb_c,gtdb_o, gtdb_f, f_spor) %>% 
  left_join(d_meta_clean,.) %>% 
  # assign non Firmicutes as non sporulators
  mutate(f_spor = if_else(str_detect(gtdb_p, "Firmicutes"), f_spor, FALSE))
```


## TaxIDs absent in GTDB data

Inspect the taxonomy we have for the hosts of these viruses. 
```{r}
d.vir %>% 
  filter(host.tax.id %in% taxid_absent) %>% 
  group_by(phylum) %>% 
  summarise(n=n())
  
```

Any non-Firmicutes we can assign as non-sporulators.

```{r}
to_add <- d.vir %>% 
  filter(host.tax.id %in% taxid_absent) %>% 
  filter(phylum != "Firmicutes") %>% 
  filter(!is.na(phylum)) %>% 
  pull(host.tax.id)

d_meta_clean <- 
  tibble(ncbi_taxid = to_add,
         f_spor = FALSE) %>% 
  bind_rows(d_meta_clean, . )
```


We are left with 42 refseq viruses that infect Firmicutes hosts and which we have not yet assigned sporulation status. Lets have a look.

```{r}
left_ID <- d.vir %>% 
  filter(host.tax.id %in% taxid_absent) %>% 
  filter(phylum == "Firmicutes") %>% 
  filter(!is.na(phylum)) %>% 
  pull(host.tax.id)

d.left <- d.vir %>% 
  filter(host.tax.id %in% left_ID) %>% 
  select(virus.name, host.tax.id ,host.name, family, genus.etc)

d.left

unique(d.left$family)
```
The remaining phages infect hosts of eight firmicutes families. Let's check if these families are classified in our gtdb list

```{r}
fam_spor %>% 
  filter(str_detect(gtdb_f,
                    regex(paste(unique(d.left$family), collapse = "|")))) %>%
  select(gtdb_f, f_spor)
```



```{r}
d.left <-
  tibble(family= unique(d.left$family)) %>% 
  mutate(gtdb_f = str_c("f__", family )) %>% 
  left_join(., fam_spor %>% select(gtdb_f, f_spor),  by = "gtdb_f") %>% 
  left_join(d.left,.,  by = "family")

d.left
```

Matching to curated GTDB family list we can classify sporulation status for all phages left, but one: *Weissella* phage WCP30 infecting a host of the family *Leuconostocaceae*. This family has been separate from, and (incorporated back)[https://www.microbiologyresearch.org/content/journal/ijsem/10.1099/ijsem.0.004107] into the *Lactobacillaceae.* Like the *Lactobacillaceae*, they do not form spores.

```{r}
d.left <- d.left %>% 
  mutate(f_spor = if_else(family == "Leuconostocaceae", FALSE, f_spor))
```

Incorporate classification into master list

```{r}
d_meta_clean <- 
  d.left %>% 
  select(ncbi_taxid = host.tax.id, gtdb_f, f_spor) %>% 
  bind_rows(d_meta_clean,.)
```

```{r}

table(d.vir$host.tax.id %in% d_meta_clean$ncbi_taxid )

d.vir %>% 
  filter(! host.tax.id %in% d_meta_clean$ncbi_taxid) %>% view
```

We have classified sporulation for the hosts of all phages that have hosts assinged. Thirty nine phages that do not have hosts are left.

```{r}
d.vir_spor <- d_meta_clean %>% 
  select(host.tax.id = ncbi_taxid , spor_host = f_spor) %>% 
  distinct() %>% 
  left_join(d.vir, ., by = "host.tax.id") 

write_csv(d.vir_spor, here("enrichment/data/Viruses/refseq_phages_wHost_spor.csv"))
  
table(d.vir_spor$spor_host)
```
These are the numbers of phages by sporulating host.
