---
title: "Formatting spore gene list for DRAM"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: inline
---
```{r setup, echo=FALSE}
library(here)
library(tidyverse)
```

# Here I combine the lists of sporulation genes from B. subtilis and C. difficle 

Here is the format description I got in a mail from Mike Shaffer:  

> In DRAM we try to add a hierarchy to metabolism so that we can categorize it and present it in an easier to understand way for DRAM users.  
The base level of this hierarchy is a single function. This function is associated with one  more identifiers from any of the databases that DRAM uses (KEGG KOs, UniRef9or0, Pfam, CAZy, MEROPS). I would caution against using UniRef90 since those categories change per release and therefore are not as stable as the other databases.  
The next level up is a module and then the subsequent levels are subheader, header and sheet. Sheet is pretty strictly defined by DRAM (for now I think we would put sporulation under MISC) but the rest can be called whatever we think is best.  
In the previous email I defined how I think the subtilwiki categories would fit into this system. We would like to put the sporulation genes in this same hierarchy so we can identify the genes in various bacteria and viruses. As an example I added the Class I Spore Coat proteins from subtiwiki to an example file with a part of the DRAM distillate. We will need to fill this out for all sporulation proteins we would like to detect.   

As an example I got this table:

```{r, message=FALSE}
dram.eg <- read_csv(here("spor_gene_list/data","DRAM_spore_example.csv" ))
dram.eg
```

## Sporulation genes  
B. subtilis list of sporulation genes in dram format:  
```{r, message=FALSE}
dram_bsub <- read_csv(here("spor_gene_list/data","dram_spore_genes_RS.csv"))

```



```{r}
write_csv(d.dram, here("spor_gene_list/data","dram_spore_genes_RS.csv"))
```

